---
title: "Lukowski: Adult human retina"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Data acquisition
Single-cell RNA-seq data from the adult human retina corresponding to [Lukowski et al 2019](https://pubmed.ncbi.nlm.nih.gov/31436334/) were retrieved from [ArrayExpress](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-7316/).

A csv file mapping cell barcodes to clusters was additionally supplied by Sam Lukowski through personal communication in October 2019. 

# Analysis of IMPG1 and IMPG2 expression in retinal cell types

## Load libraries
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(cowplot))
```

## Import metadata
```{r, echo=F}
meta <- read.table("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/CopiedFromLabComp/Lukowski_human/retina_wong_cellbc_cellid.csv",sep=",",header=T,row.names=1)
colnames(meta) <- c("CellBC","Cluster","ClusterID","CCA Cluster","CCA ClusterID")
rownames(meta) <- gsub("-",".",rownames(meta))
```

```{r, eval=F}
meta <- read.table("retina_wong_cellbc_cellid.csv",sep=",",header=T,row.names=1)
colnames(meta) <- c("CellBC","Cluster","ClusterID","CCA Cluster","CCA ClusterID")
rownames(meta) <- gsub("-",".",rownames(meta))
```

## Filter expression matrix for IMPG1 and IMPG2 expression
```{bash, eval=F}
grep -E 'GENE|IMPG1|IMPG2' ae_exp_proc_all.tsv > ae_exp_proc_all_IMPG1_IMPG2.tsv
```

## Import expression data and establish cell categories
Note categories are modeled after Fig. 1C in the paper, where clusters are summarized by broader categories

```{r,echo=F}
human_all <- as.data.frame(t(read.table("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/CopiedFromLabComp/Lukowski_human/ae_exp_proc_all_IMPG1_IMPG2.tsv",header=T,sep="\t",row.names=1,check.names=F)))
rownames(human_all) <- gsub("-",".",rownames(human_all))
```
```{r,eval=F}
human_all <- as.data.frame(t(read.table("ae_exp_proc_all_IMPG1_IMPG2.tsv",header=T,sep="\t",row.names=1,check.names=F)))
rownames(human_all) <- gsub("-",".",rownames(human_all))
```
```{r}
cat <- meta[rownames(human_all),2]
cat <- gsub("Rod PR.+","Rods",cat)
cat <- gsub("Cone PR.+","Cones",cat)
cat <- gsub("Bipolar.+","Bipolar cells",cat)
cat <- gsub("MG.+","Muller glia",cat)
cat <- gsub("Others.+","Others",cat)
cat <- gsub("Microglia.+","Microglia",cat)
cat <- gsub("RGC.+","Retinal ganglion cells",cat)
cat <- gsub("Horizontal.+","Horizontal cells",cat)
cat <- gsub("Astrocytes.+","Astrocytes",cat)
cat <- gsub("Amacrine.+","Amacrine cells",cat)
human_all$Category <- cat

hum.tbl <- as_tibble(rownames_to_column(human_all,var="Cell"))
```

## Plot boxplots of IMPG1 and IMPG2 expression
```{r}
p1 <- hum.tbl %>%
	group_by(Category) %>%
	filter(Category!="Others") %>%
	ggplot(aes(x=reorder(Category,-IMPG2),y=IMPG1)) +
	geom_boxplot(outlier.shape = NA) +
	theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),legend.position = c(0.7, 0.8)) +
	xlab("Human retina cell types") +
	ylab("IMPG1 expression")

p2 <- hum.tbl %>%
	group_by(Category) %>%
	filter(Category!="Others") %>%
	ggplot(aes(x=reorder(Category,-IMPG2),y=IMPG2)) +
	geom_boxplot(outlier.shape = NA) +
	theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),legend.position = c(0.7, 0.8)) +
	xlab("Human retina cell types") +
	ylab("IMPG2 expression")

plot_grid(p1,p2)
```
