---
title: "Peng: 3-9yo macaque retina"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Data acquisition
Single-cell RNA-seq data from the 3-9yo macaque retina corresponding to [Peng et al 2019](https://pubmed.ncbi.nlm.nih.gov/30712875/) were retrieved from the [Broad single-cell portal](https://singlecell.broadinstitute.org/single_cell/study/SCP212/molecular-specification-of-retinal-cell-types-underlying-central-and-peripheral-vision-in-primates#).
Below we additionally annotate the cluster numbers into named cell types based on the information provided in the supplied metadata file

Data were downloaded as follows
```{r,echo=FALSE,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```
```{bash, eval=F}
curl "https://singlecell.broadinstitute.org/single_cell/api/v1/bulk_download/generate_curl_config?accessions=SCP212&auth_code=Z2Yow3ap&directory=all&context=study"  -o cfg.txt; curl -K cfg.txt && rm cfg.$
```

# Analysis of IMPG1 and IMPG2 expression in retinal cell types

## Filter expression matrices for IMPG1 and IMPG2 expression
```{bash, eval=F}
grep -E 'GENE|IMPG1|IMPG2' Macaque_fov_AC_expression2.txt > Macaque_fov_AC_expression2_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_fov_BC_expression.txt > Macaque_fov_BC_expression_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_fov_EpiImmune_expression.txt > Macaque_fov_EpiImmune_expression_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_fov_HC_expression.txt > Macaque_fov_HC_expression_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_fov_PR_expression.txt > Macaque_fov_PR_expression_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_fov_RGC_expression.txt > Macaque_fov_RGC_expression_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_per_AC_expression.txt > Macaque_per_AC_expression_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_per_BC_expression2.txt > Macaque_per_BC_expression2_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_per_EpiImmune_expression.txt > Macaque_per_EpiImmune_expression_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_per_HC_expression.txt > Macaque_per_HC_expression_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_per_PR_expression.txt > Macaque_per_PR_expression_IMPG1_IMPG2.txt
grep -E 'GENE|IMPG1|IMPG2' Macaque_per_RGC_expression.txt > Macaque_per_RGC_expression_IMPG1_IMPG2.txt
```

## Load libraries
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(cowplot))
```
## Import metadata
Categories are modeled after Fig. 1C in the paper, where clusters are summarized by broader categories

```{r,echo=F,message=FALSE,warning=FALSE}
meta <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_NN_RGC_AC_BC_HC_PR_metadata_3.txt", col_names=c("CellID","Cluster","Subcluster"), skip=2) %>%
        mutate(CellID = str_replace_all(CellID,"-","."))
```

```{r,eval=F}
meta <- read_csv("Macaque_NN_RGC_AC_BC_HC_PR_metadata_3.txt", col_names=c("CellID","Cluster","Subcluster"), skip=2) %>%
	mutate(CellID = str_replace_all(CellID,"-","."))
```

## Import expression data
```{r,echo=F,message=FALSE,warning=FALSE}
Macaque_fov_AC <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_fov_AC_expression2_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Amacrine cells")

Macaque_fov_BC <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_fov_BC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Bipolar cells")
	
Macaque_fov_EpiImmune <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_fov_EpiImmune_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%	
	mutate(Category = str_replace_all(Cluster, c("MG"="Muller glia", "Mic"="Microglia", "Endo"="Endothelial")))

Macaque_fov_HC <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_fov_HC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Horizontal cells")

Macaque_fov_PR <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_fov_PR_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	mutate(Category = str_replace_all(Cluster,"M/L Cones|S Cones", "Cones"))

Macaque_fov_RGC <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_fov_RGC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Retinal ganglion cells")


Macaque_per_AC <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_per_AC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Amacrine cells")

Macaque_per_BC <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_per_BC_expression2_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Bipolar cells")
	
Macaque_per_EpiImmune <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_per_EpiImmune_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%	
	mutate(Category = str_replace_all(Cluster, c("MG"="Muller glia", "Mic"="Microglia", "Endo"="Endothelial")))

Macaque_per_HC <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_per_HC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Horizontal cells")

Macaque_per_PR <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_per_PR_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	mutate(Category = str_replace_all(Cluster,"M/L Cones|S Cones", "Cones"))

Macaque_per_RGC <- read_csv("/proj/jmsimon/Philpot/singleCellRetinaAnalysis/Macaque_per_RGC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Retinal ganglion cells")

```


```{r,eval=F}
Macaque_fov_AC <- read_csv("Macaque_fov_AC_expression2_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Amacrine cells")

Macaque_fov_BC <- read_csv("Macaque_fov_BC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Bipolar cells")
	
Macaque_fov_EpiImmune <- read_csv("Macaque_fov_EpiImmune_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%	
	mutate(Category = str_replace_all(Cluster, c("MG"="Muller glia", "Mic"="Microglia", "Endo"="Endothelial")))

Macaque_fov_HC <- read_csv("Macaque_fov_HC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Horizontal cells")

Macaque_fov_PR <- read_csv("Macaque_fov_PR_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	mutate(Category = str_replace_all(Cluster,"M/L Cones|S Cones", "Cones"))

Macaque_fov_RGC <- read_csv("Macaque_fov_RGC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Retinal ganglion cells")


Macaque_per_AC <- read_csv("Macaque_per_AC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Amacrine cells")

Macaque_per_BC <- read_csv("Macaque_per_BC_expression2_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Bipolar cells")
	
Macaque_per_EpiImmune <- read_csv("Macaque_per_EpiImmune_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%	
	mutate(Category = str_replace_all(Cluster, c("MG"="Muller glia", "Mic"="Microglia", "Endo"="Endothelial")))

Macaque_per_HC <- read_csv("Macaque_per_HC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Horizontal cells")

Macaque_per_PR <- read_csv("Macaque_per_PR_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	mutate(Category = str_replace_all(Cluster,"M/L Cones|S Cones", "Cones"))

Macaque_per_RGC <- read_csv("Macaque_per_RGC_expression_IMPG1_IMPG2.txt", name_repair="minimal") %>%
	pivot_longer(cols=!GENE,names_to="CellID",values_to="Expression") %>%
	pivot_wider(names_from=GENE,values_from=Expression) %>%
	mutate(CellID = str_replace_all(CellID,"-",".")) %>%
	inner_join(meta,by="CellID") %>%
	add_column(Category = "Retinal ganglion cells")
```

```{r}
macaque_all <- bind_rows(Macaque_fov_AC,Macaque_fov_BC,Macaque_fov_EpiImmune,Macaque_fov_HC,Macaque_fov_PR,Macaque_fov_RGC,Macaque_per_AC,Macaque_per_BC,Macaque_per_EpiImmune,Macaque_per_HC,Macaque_per_PR,Macaque_per_RGC)
```

## Plot boxplots of IMPG1 and IMPG2 expression
Split and color fovea and periphery separately

```{r}
p1 <- macaque_all %>%
	group_by(Category) %>%
	ggplot(aes(x=reorder(Category,-IMPG2),y=IMPG1,fill=Subcluster)) +
	geom_boxplot(outlier.shape = NA) +
	theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),legend.position = c(0.7, 0.8)) +
	xlab("Macaque retina cell types") +
	ylab("IMPG1 expression")

p2 <- macaque_all %>%
	group_by(Category) %>%
	ggplot(aes(x=reorder(Category,-IMPG2),y=IMPG2,fill=Subcluster)) +
	geom_boxplot(outlier.shape = NA) +
	theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),legend.position = c(0.7, 0.8)) +
	xlab("Macaque retina cell types") +
	ylab("IMPG2 expression")

plot_grid(p1,p2)
```
